---
title: "Ethiopia Flood Susceptibility"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
  eval: false
  results: "asis"
  out.width: "100%"
editor: visual
project:
     execute-dir: project
---

## Introduction

The seasons covered under this are:

-   **Apr-May-Jun**: Coinciding with the Gu/Ganna rainy season in Somali and Oromia regions. This period overlaps with the end of the Sugum rains in Afar region and the Belg rains.

-   **Annual**: This looks at the entire year, and is expected to be the highest value overall.

-   **Jun-Jul-Aug-Sep**: This produces flood exposure estimates based on the Kiremt rainy season. It also overlaps with the Karma rains in the Afar region and the Karran rains in Somali region.

-   **Mar-Apr-May**: This looks at the Belg rainy season and overlaps with the beginning of the Gu/Ganna rains in Somali and Oromia as well as the Sugum rains in Afar region.

-   **Oct-Nov-Dec**: This looks at the Deyr rains in Somali region, Hageya rains in Oromia region and part of the Hageya/Hagayita rains in Segen and rains in South Omo.

This analysis uses FloodScan (1998-2022) & WorldPop (2022) to generate flood estimates for Ethiopia. Analysis is performed at regional & zonal level.

```{r ,eval=T, echo=F, results='hide'}
library(targets)
library(tidyverse)
library(here)
library(scales)
library(sf)
library(gt)
library(gghdx)
library(glue)
gghdx()

```

```{r ,eval=T}

# Load  a bunch of targets so it's easy to make visuals
tar_load(lgt_adm1_bin_method,store = here("flood_target_store"))
tar_load(lgt_adm2_bin_method,store = here("flood_target_store"))
tar_load(lgt_adm3_bin_method,store = here("flood_target_store"))
tar_load(lgdf_adm,store = here("flood_target_store"))
tar_load(df_adm1_mean_stat,store = here("flood_target_store"))
tar_load(df_adm2_mean_stat,store = here("flood_target_store"))
tar_load(df_adm3_mean_stat,store = here("flood_target_store"))
tar_load(df_adm2_stats_bin_method,store = here("flood_target_store"))
tar_load(df_adm3_stats_bin_method,store = here("flood_target_store"))
```

```{r, eval=TRUE}
lgt_adm1_bin_method <- df_adm1_mean_stat %>%
      split(.$adm0_en) %>%
      imap(\(dft, country_name){
        dft %>%
          dplyr::filter(thresh == 0.005) %>%
          dplyr::select(adm0_en, adm1_en, season, 
                        mean_exposed, total_pop, pct_exposed) %>%
          gt() %>%
          cols_hide(columns = "adm0_en") %>%
          cols_label(
            # adm0_en = "Country",
            adm1_en = "Region",
            season = "Time Period",
            mean_exposed = "Average population exposed",
            total_pop = "Total population",
            pct_exposed = "Percent population exposed"
          ) %>%
          fmt_number(
            columns = c(
              "mean_exposed",
              "total_pop",
              "pct_exposed"
            ),
            decimals = 0, n_sigfig = 3
          ) %>%
          fmt_percent(columns = pct_exposed, decimals = 0) %>%
          data_color(
            columns = c(
              "mean_exposed",
              "total_pop",
              "pct_exposed"
            ),
            method = "numeric",
            palette = "YlOrRd",
            reverse = F
          ) %>%
          tab_options(
            row_group.as_column = TRUE
          ) %>%
          tab_header(
            title = paste0("Approximate population susceptible to flooding", country_name)
          )
      })
lgt_adm2_bin_method <- df_adm2_mean_stat %>%
      split(.$adm0_en) %>%
      imap(\(dft, country_name){
        dft %>%
          dplyr::filter(thresh == 0.005) %>%
          dplyr::select(adm0_en, adm1_en, adm2_en, season, mean_exposed, total_pop, pct_exposed) %>%
          group_by(adm1_en) %>%
          gt() %>%
          cols_hide(columns = "adm0_en") %>%
          cols_label(
            # adm0_en = "Country",
            adm1_en = "Region",
            adm2_en = "District",
            season = "Time Period",
            mean_exposed = "Average population exposed",
            total_pop = "Total population",
            pct_exposed = "Percent population exposed"
          ) %>%
          fmt_number(
            columns = c(
              "season",
              "mean_exposed",
              "total_pop",
              "pct_exposed"
            ),
            decimals = 0, n_sigfig = 3
          ) %>%
          fmt_percent(columns = pct_exposed, decimals = 0) %>%
          data_color(
            columns = c(
              "mean_exposed",
              "total_pop",
              "pct_exposed"
            ),
            method = "numeric",
            palette = "YlOrRd",
            reverse = F
          ) %>%
          tab_options(
            row_group.as_column = TRUE
          ) %>%
          tab_header(
            title = paste0("Approximate population susceptible to flooding", country_name)
          )
      })


```

## By Region

:::: panel-tabset
## **Method: Flooded Fraction multiplied by the population**

::: panel-tabset
## **Ethiopia**

```{r ,eval=T}
lgt_adm1_bin_method$Ethiopia
```

## **Map (Population)**

Population in regions potentially exposed

```{r, eval=T, warning=FALSE}
gdf_adm1_stat<- lgdf_adm$adm1 %>% 
  filter(adm0_pcode == "ET") %>%
  left_join(df_adm1_mean_stat %>%
              dplyr::filter(thresh == 0.005), 
            by = "adm1_pcode")

gdf_adm1_stat %>% 
  ggplot()+
  geom_sf(aes(fill=mean_exposed)) +
  scale_fill_gradient(
    trans="pseudo_log",
    low="white",
    # midpoint="white",
    
    high="red",
    
    breaks= c(1e2,1e3,1e4,1e5,1e6,1e7,1e8,1e9,1e11),
    labels=scales::label_comma()) +
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1) +
  facet_wrap(~season) +
  guides(fill = guide_legend(title = "Mean Pop Exposed")) + 
  # geom_sf_text(data = lgdf_adm$adm1 %>% filter(adm0_pcode == "ET"), aes(label = adm1_en), color = "black", check_overlap = T) +
  # scale_fill_viridis_c()+
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Population Exposed"
  )
```

## **Map (Percentage of Population)**

Percent of population in regions potentially exposed

```{r, eval=T, warning=FALSE}
gdf_adm1_stat<- lgdf_adm$adm1 %>% 
  dplyr::filter(adm0_pcode == "ET") %>%
  left_join(df_adm1_mean_stat %>%
              dplyr::filter(thresh == 0.005), 
            by = "adm1_pcode")

gdf_adm1_stat %>% 
  ggplot()+
  geom_sf(aes(fill=pct_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    high="red",
    breaks= c(seq(0, 0.8, by=0.02)),
    # breaks= c(seq(0,0.2,by=0.04)),
    labels=scales::label_percent()
  )+
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1)+
  facet_wrap(~season) +
  #geom_sf_text(data = lgdf_adm$adm1 %>% filter(adm0_pcode == "ET"), aes(label = adm1_en), color = "black", check_overlap = T) +
  # scale_fill_viridis_c()+
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Percentage Exposed"
  )
```
:::
::::

## By Zone

:::: panel-tabset
## **Method: Flooded Fraction multiplied by the population**

::: panel-tabset
## **Ethiopia**

```{r ,eval=T}
lgt_adm2_bin_method$Ethiopia
```

## **Map (Population)**

Population in zones potentially exposed

```{r, eval=T, warning=FALSE}
gdf_adm2_stat<- lgdf_adm$adm2 %>% 
  dplyr::filter(adm0_pcode == "ET") %>%
  left_join(df_adm2_mean_stat %>%
              dplyr::filter(thresh == 0.005), by = "adm2_pcode")

gdf_adm2_stat %>% 
  ggplot()+
  geom_sf(aes(fill=mean_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    high="red",
    breaks= c(seq(0, 1000000, 50000)),
    labels=scales::label_comma()
  )+
  geom_sf(data= lgdf_adm$adm1 %>% 
            dplyr::filter(adm0_pcode == "ET") ,
          fill=NA,color="black",lwd=0.5)+
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1)+
  facet_wrap(~season) +
  # scale_fill_viridis_c()+
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Population Exposed"
  )
```

## **Map (Percentage of Population)**

Percent of population in zones potentially exposed

```{r, eval=T}

gdf_adm2_stat %>% 
  ggplot()+
  geom_sf(aes(fill=pct_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    high="red",
    breaks= c(seq(0,0.8,by=0.05)),
    # breaks= c(seq(0,0.2,by=0.04)),
    labels=scales::label_percent()
  )+
  geom_sf(data= lgdf_adm$adm1 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=0.5)+
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1)+
  facet_wrap(~season) +
  # scale_fill_viridis_c()+
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Percentage Exposed"
  )
```
:::
::::

## By Woreda

:::: panel-tabset
## **Method: Flooded Fraction multiplied by the population**

::: panel-tabset
## **Ethiopia**

```{r ,eval=T}
lgt_adm3_bin_method$Ethiopia
```

## **Map (Population)**

Population in zones potentially exposed

```{r, eval=T, warning=FALSE}
gdf_adm3_stat <- lgdf_adm$adm3 %>%    
  dplyr::filter(adm0_pcode == "ET") %>%   
  left_join(df_adm3_mean_stat %>%               
              dplyr::filter(thresh == 0.005), 
            by = "adm3_pcode") 
gdf_adm3_stat %>%    
  ggplot()+   
  geom_sf(aes(fill=mean_exposed))+   
  scale_fill_gradient(     
    # trans="pseudo_log",
    low="white",     
    # midpoint="white",     
    high="red",     
    breaks= c(seq(0, 1000000, 50000)),     
    labels=scales::label_comma())+   
  geom_sf(data= lgdf_adm$adm1 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=0.5)+   
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1)+   
  facet_wrap(~season) +   
  # scale_fill_viridis_c()+   
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Population Exposed"
  )
```

## **Map (Percentage of Population)**

Percent of population in zones potentially exposed

```{r, eval=T}
gdf_adm3_stat %>%    
  ggplot()+   
  geom_sf(aes(fill=pct_exposed))+   
  scale_fill_gradient(     
    # trans="pseudo_log",     
    low="white",     
    # midpoint="white",     
    high="red",          
    breaks= c(seq(0,0.8,by=0.05)),     
    # breaks= c(seq(0,0.2,by=0.04)),     
    labels=scales::label_percent())+   
  geom_sf(data= lgdf_adm$adm1 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=0.5)+  
  geom_sf(data= lgdf_adm$adm0 %>% 
            dplyr::filter(adm0_pcode == "ET"),
          fill=NA,color="black",lwd=1)+   
  facet_wrap(~season) +   
  # scale_fill_viridis_c()+   
  theme_void() +
  theme(
    strip.text = element_text(size = 18, face = "bold") 
  ) +
  labs(
    fill = "Percentage Exposed"
  )
```
:::
::::

## Method Note

Yearly maximum composite flood composite rasters were made for each year of available FloodScan Data (1998-2022). The seasonal maximum composites were then re-sampled to the [World Pop 2020 UN Adjusted (1 km) Population Count Grid](https://hub.worldpop.org/). Flood fraction was then combined with world population following the method described below:

**Method** A minimum flood fraction threshold was used and it is assumed that flood fractions above this are "flooded". The number of people exposed to flooding is calculated as the population beneath any "flooded" pixel multiplied by the flood fraction.
