---
title: "Ethiopia Flood Susceptibility"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
  eval: false
  results: "asis"
  out.width: "100%"
editor: visual
project:
     execute-dir: project
---

## Introduction

This analysis uses FloodScan (1998-2022) & WorldPop (2022) to generate annual flood estimates for Ethiopia.

-   Analysis is performed at admin 1 & admin 2

```{r ,eval=T, echo=F, results='hide'}
library(targets)
library(tidyverse)
library(here)
library(scales)
library(sf)
library(gghdx)
library(glue)
gghdx()

```

```{r ,eval=T}

# Load  a bunch of targets so it's easy to make visuals
tar_load(lgt_adm1_bin_method_annual,store = here("flood_target_store"))
tar_load(lgt_adm2_bin_method_annual,store = here("flood_target_store"))
tar_load(lgdf_adm,store = here("flood_target_store"))
tar_load(df_adm1_mean_stat,store = here("flood_target_store"))
tar_load(df_adm2_mean_stat,store = here("flood_target_store"))
tar_load(df_adm2_stats_bin_method,store = here("flood_target_store"))
```

## Admin 1

::: panel-tabset
## Method: Flooded Fraction is 20% and above

::: panel-tabset
## Ethiopia

```{r ,eval=T}
lgt_adm1_bin_method_annual$Ethiopia
```

## Map (Absolute)

```{r, eval=T, warning=FALSE}
gdf_adm1_stat<- lgdf_adm$adm1 %>% 
  filter(adm0_pcode == "ET") %>%
  left_join(df_adm1_mean_stat %>%
              filter(season == "Annual"))

gdf_adm1_stat %>% 
  ggplot()+
  geom_sf(aes(fill=mean_exposed))+
  scale_fill_gradient(
    trans="pseudo_log",
    low="white",
    # midpoint="white",
    
    high="red",
    
    breaks= c(1e3,1e4,1e5,1e6,1e7,1e8,1e9,1e11),
    labels=scales::label_comma()
  )+
  geom_sf(data= lgdf_adm$adm0 %>% filter(adm0_pcode == "ET"),fill=NA,color="black",lwd=1) +
  geom_sf_text(data = lgdf_adm$adm1 %>% filter(adm0_pcode == "ET"), aes(label = adm1_en), color = "black", check_overlap = T) +
  # scale_fill_viridis_c()+
  theme_void()
```

## Map (%)

\% district potentially exposed

```{r, eval=T, warning=FALSE}
gdf_adm1_stat<- lgdf_adm$adm1 %>% 
  filter(adm0_pcode == "ET") %>%
  left_join(df_adm1_mean_stat %>%
              filter(season == "Annual"))

gdf_adm1_stat %>% 
  ggplot()+
  geom_sf(aes(fill=pct_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    
    high="red",
    
    breaks= c(seq(0,0.1,by=0.02),seq(0.15,0.25, by=0.05)),
    # breaks= c(seq(0,0.2,by=0.04)),
    labels=scales::label_percent()
  )+
  geom_sf(data= lgdf_adm$adm0 %>% filter(adm0_pcode == "ET"),fill=NA,color="black",lwd=1)+
  geom_sf_text(data = lgdf_adm$adm1 %>% filter(adm0_pcode == "ET"), aes(label = adm1_en), color = "black", check_overlap = T) +
  # scale_fill_viridis_c()+
  theme_void()
```
:::
:::

## Admin 2

::: panel-tabset
## Method 1

::: panel-tabset
## Ethiopia

```{r ,eval=T}
lgt_adm2_bin_method_annual$Ethiopia
```

## Map (Absolute)

```{r, eval=T, warning=FALSE}
gdf_adm2_stat<- lgdf_adm$adm2 %>% 
  filter(adm0_pcode == "ET") %>%
  left_join(df_adm2_mean_stat %>%
              filter(season == "Annual"))

gdf_adm2_stat %>% 
  ggplot()+
  geom_sf(aes(fill=mean_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    high="red",
    breaks= c(10,50,100,10000,50000,75000,100000, 200000),
    labels=scales::label_comma()
  )+
  geom_sf(data= lgdf_adm$adm1 %>% filter(adm0_pcode == "ET") ,fill=NA,color="black",lwd=0.5)+
  geom_sf(data= lgdf_adm$adm0 %>% filter(adm0_pcode == "ET"),fill=NA,color="black",lwd=1)+
  # scale_fill_viridis_c()+
  theme_void()
```

## Map (%)

\% district potentially exposed

```{r, eval=T}

gdf_adm2_stat %>% 
  ggplot()+
  geom_sf(aes(fill=pct_exposed))+
  scale_fill_gradient(
    # trans="pseudo_log",
    low="white",
    # midpoint="white",
    high="red",
    
    breaks= c(seq(0,0.1,by=0.02),seq(0.15,0.25, by=0.05)),
    # breaks= c(seq(0,0.2,by=0.04)),
    labels=scales::label_percent()
  )+
  geom_sf(data= lgdf_adm$adm1 %>% filter(adm0_pcode == "ET"),fill=NA,color="black",lwd=0.5)+
  geom_sf(data= lgdf_adm$adm0 %>% filter(adm0_pcode == "ET"),fill=NA,color="black",lwd=1)+
  # scale_fill_viridis_c()+
  theme_void()
```
:::
:::

## Method Note

Yearly maximum composite flood composite rasters were made for each year of available FloodScan Data (1998-2022). The yearly maximum composites were then re-sampled to the [World Pop 2020 UN Adjusted (1 km) Population Count Grid](https://hub.worldpop.org/). Flood fraction was then combined with world population folllowing 2 methods to get population exposure estimates:

**Method 1** All flood fractions \> 20 % were assumed to be "flooded". The entire population beneath any "flooded" pixel was then considered flooded
