---
title: "Mozambique January 2026 Flood Event"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    css: custom.css
    embed-resources: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(cumulus)
library(DBI)
library(zoo)
library(dbplyr)

con <- pg_con()
df_lookup <- cumulus::blob_load_admin_lookup()

provinces <- c("Maputo", "Sofala", "Gaza")

df_lookup_aoi <- df_lookup |>
  filter(
    ADM_LEVEL == 1,
    ISO3 == "MOZ",
    ADM1_NAME %in% provinces
  )

imerg <- tbl(con, "imerg") |>
  filter(
    adm_level == 1,
    pcode %in% df_lookup_aoi$ADM1_PCODE
  ) |>
  collect() |>
  left_join(
    df_lookup_aoi |> select(pcode = ADM1_PCODE, adm1_name = ADM1_NAME),
    by = "pcode"
  ) |>
  select(iso3, pcode, adm1_name, valid_date, mean) |>
  arrange(adm1_name, valid_date)
```

# Rainfall Analysis (IMERG)

## Recent Rainfall Context

**Method:** Daily IMERG satellite precipitation aggregated to province level.

**Key finding:** Mid-January 2026 shows a distinct rainfall spike across all three provinces, with peak daily totals reaching 30-50+ mm/day.

```{r recent-rainfall}
#| fig-height: 7

imerg_recent <- imerg |>
  filter(valid_date >= max(valid_date) - 60)

top_3_days <- imerg_recent |>
  group_by(adm1_name) |>
  slice_max(mean, n = 3) |>
  ungroup()

ggplot(imerg_recent, aes(x = valid_date, y = mean, color = adm1_name)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1) +
  geom_point(data = top_3_days, size = 2.5, shape = 21, fill = "white", stroke = 1.2) +
  geom_text(
    data = top_3_days,
    aes(label = format(valid_date, "%b %d")),
    vjust = -1.2, size = 2.5, show.legend = FALSE
  ) +
  facet_wrap(~adm1_name, ncol = 1) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  labs(
    title = "Daily Precipitation - Last 60 Days",
    subtitle = "Top 3 days labeled per province",
    x = NULL,
    y = "Precipitation (mm/day)",
    caption = "Data: IMERG"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
```

## January 2026 Accumulations

**Method:** Rolling 3-day and 7-day precipitation sums identify peak accumulation windows.

**Key finding:** Maximum accumulations occurred mid-January. Shaded regions show the worst 3-day and 7-day periods per province.

```{r rolling-calcs}
imerg_rolling <- imerg |>
  group_by(adm1_name) |>
  mutate(
    roll3_sum = rollsum(mean, k = 3, fill = NA, align = "right"),
    roll7_sum = rollsum(mean, k = 7, fill = NA, align = "right")
  ) |>
  ungroup()

jan_2026 <- imerg_rolling |>
  filter(year(valid_date) == 2026, month(valid_date) == 1)

max_3day <- jan_2026 |>
  filter(!is.na(roll3_sum)) |>
  group_by(adm1_name) |>
  slice_max(roll3_sum, n = 1) |>
  mutate(
    start_date = valid_date - 2,
    period_label = paste0(format(start_date, "%b %d"), "-", format(valid_date, "%d"))
  ) |>
  ungroup()

max_7day <- jan_2026 |>
  filter(!is.na(roll7_sum)) |>
  group_by(adm1_name) |>
  slice_max(roll7_sum, n = 1) |>
  mutate(
    start_date = valid_date - 6,
    period_label = paste0(format(start_date, "%b %d"), "-", format(valid_date, "%d"))
  ) |>
  ungroup()
```

```{r jan-timeseries}
#| fig-height: 7

ggplot(jan_2026, aes(x = valid_date, y = mean)) +
  geom_rect(
    data = max_7day,
    aes(xmin = start_date, xmax = valid_date, ymin = -Inf, ymax = Inf),
    fill = "steelblue", alpha = 0.15, inherit.aes = FALSE
  ) +
  geom_rect(
    data = max_3day,
    aes(xmin = start_date, xmax = valid_date, ymin = -Inf, ymax = Inf),
    fill = "steelblue", alpha = 0.25, inherit.aes = FALSE
  ) +
  geom_col(fill = "steelblue", width = 0.8) +
  geom_label(
    data = max_3day,
    aes(x = start_date + 1, y = Inf,
        label = paste0("3-day: ", round(roll3_sum, 0), " mm")),
    vjust = 1.2, hjust = 0, size = 3.5, fontface = "bold",
    color = "grey20", fill = "white", label.size = 0.3, label.r = unit(0.15, "lines")
  ) +
  geom_label(
    data = max_7day,
    aes(x = start_date + 3, y = Inf,
        label = paste0("7-day: ", round(roll7_sum, 0), " mm")),
    vjust = 2.8, hjust = 0, size = 3.5,
    color = "grey40", fill = "white", label.size = 0.3, label.r = unit(0.15, "lines")
  ) +
  facet_wrap(~adm1_name, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  labs(
    title = "January 2026 Daily Precipitation",
    subtitle = "Shaded: max 3-day (dark) and 7-day (light) accumulation periods",
    x = NULL,
    y = "Precipitation (mm/day)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
```

## Return Period Analysis

**Method:** Empirical return periods using Weibull plotting position. January 2026 event compared against historical annual maxima from IMERG record.

**Key finding:** The 7-day accumulation shows higher return periods than the 3-day, indicating sustained rainfall over a full week was the more exceptional aspect of this event. 

### 3-Day Sum

```{r rp-3day}
#| fig-height: 5

annual_max_3day <- imerg_rolling |>
  filter(!is.na(roll3_sum)) |>
  mutate(year = year(valid_date)) |>
  group_by(adm1_name, year) |>
  slice_max(roll3_sum, n = 1) |>
  ungroup() |>
  group_by(adm1_name) |>
  mutate(
    n_years = n(),
    rank = rank(-roll3_sum, ties.method = "first"),
    return_period = (n_years + 1) / rank
  ) |>
  ungroup()

rp_3day <- max_3day |>
  left_join(
    annual_max_3day |> select(adm1_name, annual_max = roll3_sum),
    by = "adm1_name", relationship = "many-to-many"
  ) |>
  group_by(adm1_name, start_date, valid_date, roll3_sum, period_label) |>
  summarise(
    n_exceeded = sum(roll3_sum > annual_max),
    n_years = n(),
    rp_years = (n_years + 1) / pmax(1, n_years - n_exceeded),
    .groups = "drop"
  )

top10_3day <- annual_max_3day |>
  group_by(adm1_name) |>
  slice_min(rank, n = 10) |>
  ungroup()

ggplot(annual_max_3day, aes(x = return_period, y = roll3_sum)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3, alpha = 0.5) +
  geom_text(
    data = top10_3day,
    aes(label = paste0("'", substr(year, 3, 4))),
    hjust = -0.5, vjust = 0.5, size = 2.5
  ) +
  geom_hline(
    data = rp_3day,
    aes(yintercept = roll3_sum),
    color = "#D62828", linetype = "dashed", linewidth = 1
  ) +
  geom_label(
    data = rp_3day,
    aes(x = Inf, y = roll3_sum,
        label = paste0(period_label, "\n~", round(rp_years, 1), "-yr")),
    hjust = 1.05, vjust = 0.5, size = 2.5,
    label.size = 0, fill = "white", color = "#D62828", fontface = "bold"
  ) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50)) +
  facet_wrap(~adm1_name, scales = "free_y") +
  labs(
    title = "Return Period: 3-Day Sum",
    subtitle = "January 2026 event in red",
    x = "Return Period (years)",
    y = "3-Day Sum (mm)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
```

### 7-Day Sum

```{r rp-7day}
#| fig-height: 5

annual_max_7day <- imerg_rolling |>
  filter(!is.na(roll7_sum)) |>
  mutate(year = year(valid_date)) |>
  group_by(adm1_name, year) |>
  slice_max(roll7_sum, n = 1) |>
  ungroup() |>
  group_by(adm1_name) |>
  mutate(
    n_years = n(),
    rank = rank(-roll7_sum, ties.method = "first"),
    return_period = (n_years + 1) / rank
  ) |>
  ungroup()

rp_7day <- max_7day |>
  left_join(
    annual_max_7day |> select(adm1_name, annual_max = roll7_sum),
    by = "adm1_name", relationship = "many-to-many"
  ) |>
  group_by(adm1_name, start_date, valid_date, roll7_sum, period_label) |>
  summarise(
    n_exceeded = sum(roll7_sum > annual_max),
    n_years = n(),
    rp_years = (n_years + 1) / pmax(1, n_years - n_exceeded),
    .groups = "drop"
  )

top10_7day <- annual_max_7day |>
  group_by(adm1_name) |>
  slice_min(rank, n = 10) |>
  ungroup()

ggplot(annual_max_7day, aes(x = return_period, y = roll7_sum)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3, alpha = 0.5) +
  geom_text(
    data = top10_7day,
    aes(label = paste0("'", substr(year, 3, 4))),
    hjust = -0.5, vjust = 0.5, size = 2.5
  ) +
  geom_hline(
    data = rp_7day,
    aes(yintercept = roll7_sum),
    color = "#D62828", linetype = "dashed", linewidth = 1
  ) +
  geom_label(
    data = rp_7day,
    aes(x = Inf, y = roll7_sum,
        label = paste0(period_label, "\n~", round(rp_years, 1), "-yr")),
    hjust = 1.05, vjust = 0.5, size = 2.5,
    label.size = 0, fill = "white", color = "#D62828", fontface = "bold"
  ) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50)) +
  facet_wrap(~adm1_name, scales = "free_y") +
  labs(
    title = "Return Period: 7-Day Sum",
    subtitle = "January 2026 event in red",
    x = "Return Period (years)",
    y = "7-Day Sum (mm)"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  )
```

## Summary

**Takeaway:** Table summarizes return period estimates by province. Higher return periods = rarer events.

```{r summary-table}
summary_table <- bind_rows(
  rp_3day |>
    mutate(Metric = "3-Day") |>
    select(Province = adm1_name, Metric, Period = period_label,
           `Rainfall (mm)` = roll3_sum, `Return Period (yr)` = rp_years),
  rp_7day |>
    mutate(Metric = "7-Day") |>
    select(Province = adm1_name, Metric, Period = period_label,
           `Rainfall (mm)` = roll7_sum, `Return Period (yr)` = rp_years)
) |>
  mutate(`Rainfall (mm)` = round(`Rainfall (mm)`, 1),
         `Return Period (yr)` = round(`Return Period (yr)`, 1)) |>
  arrange(Province, Metric)

knitr::kable(summary_table)
```

# Flood Exposure Analysis (FloodScan)

```{r}

df_lookup_adm2 <- df_lookup |> 
  filter(
    ADM1_PCODE %in% df_lookup_aoi$ADM1_PCODE,
    ADM_LEVEL ==2
  )

floodscan_adm2_unlabelled <- tbl(con, in_schema("app", "floodscan_exposure")) |>                           
    filter(pcode %in% df_lookup_adm2$ADM2_PCODE) |>                             
    collect()

floodscan_adm2 <- floodscan_adm2_unlabelled |> 
  left_join(
    df_lookup_adm2 |> 
      select(adm1_name=ADM1_NAME, adm2_name = ADM2_NAME, pcode = "ADM2_PCODE")
  )


```

## FloodScan Exposure by District

**Method:** FloodScan flood extent crossed with WorldPop population to estimate people exposed per district.

**Key finding:** Heatmap shows temporal evolution of flood exposure. Districts ordered by total exposure - worst-affected at top. Darkest colors = highest population exposure.

[Explore population exposure data interactively here](https://chd-ds-floodexposure-monitoring.azurewebsites.net/)

```{r floodscan-heatmap}
#| fig-height: 12
#| fig-width: 10

floodscan_recent <- floodscan_adm2 |>
  filter(valid_date >= max(valid_date) - 30) |>
  complete(
    valid_date = seq.Date(min(valid_date), max(valid_date), by = "day"),
    nesting(adm1_name, adm2_name),
    fill = list(sum = NA)
  )

ggplot(floodscan_recent, aes(x = valid_date, y = reorder(adm2_name, sum, \(x) sum(x, na.rm = TRUE)), fill = sum)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(
    option = "inferno",
    direction = -1,
    labels = scales::comma,
    name = "Flood Exposure\n(sum)"
  ) +
  scale_x_date(date_breaks = "1 day", date_labels = "%b %d", expand = c(0, 0)) +
  facet_wrap(~adm1_name, ncol = 1, scales = "free_y") +
  labs(
    title = "FloodScan Exposure by District",
    subtitle = "Last 30 days | Districts ordered by total exposure",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    legend.position = "right",
    legend.key.height = unit(1.5, "cm")
  )
```

### FloodScan Exposure Table

**Takeaway:** Same data in tabular form with exact values. Color intensity indicates exposure magnitude.

```{r floodscan-gt-table}
library(gt)

floodscan_wide <- floodscan_recent |>
  mutate(date_label = format(valid_date, "%b %d")) |>
  select(adm1_name, adm2_name, date_label, sum) |>
  pivot_wider(
    names_from = date_label,
    values_from = sum
  ) |>
  arrange(adm1_name, adm2_name)

date_cols <- floodscan_wide |>
  select(-adm1_name, -adm2_name) |>
  names()

sum_range <- range(floodscan_recent$sum, na.rm = TRUE)

floodscan_wide |>
  gt(groupname_col = "adm1_name") |>
  fmt_number(
    columns = all_of(date_cols),
    decimals = 0,
    use_seps = TRUE
  ) |>
  sub_missing(missing_text = "-") |>
  data_color(
    columns = all_of(date_cols),
    palette = "inferno",
    reverse = TRUE,
    na_color = "white",
    domain = sum_range
  ) |>
  cols_label(adm2_name = "District") |>
  tab_header(
    title = "FloodScan Exposure by District",
    subtitle = "Daily values (last 30 days)"
  ) |>
  tab_style(
    style = cell_text(size = px(10)),
    locations = cells_body()
  ) |>
  tab_style(
    style = cell_text(size = px(9)),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_options(
    table.font.size = px(11),
    row_group.background.color = "#f0f0f0",
    column_labels.font.weight = "bold"
  )
```

## Flood Exposure Maps

**Method:** District-level population exposure mapped for the last 3 days. Gaza, Maputo, and Sofala provinces outlined in red.

**Key finding:** Top 3 most-affected districts labeled per province. Exposure concentrated in southern Mozambique, particularly along river valleys and low-lying coastal areas.

```{r floodscan-choropleth}
#| fig-height: 36
#| fig-width: 14

library(sf)
library(ggrepel)

sf_moz <- cumulus::download_fieldmaps_sf(iso3 = "moz", layer = "moz_adm2")

# Last 3 days of exposure data
floodscan_last3 <- floodscan_adm2 |>
  filter(valid_date >= max(valid_date) - 2)

# Join exposure to spatial (all provinces), filter out NA dates
sf_exposure <- sf_moz$moz_adm2 |>
  left_join(floodscan_last3, by = c("ADM2_PCODE" = "pcode")) |>
  filter(!is.na(valid_date))

# All province boundaries (light)
sf_provinces <- sf_moz$moz_adm2 |>
  group_by(ADM1_PT) |>
  summarise(geom = st_union(geom))

# Highlight boundaries for worst-affected provinces
sf_highlight <- sf_provinces |>
  filter(ADM1_PT %in% c("Gaza", "Maputo", "Sofala"))

# Province label positions (centroids)
sf_prov_labels <- sf_highlight |>
  mutate(
    centroid = st_centroid(geom),
    x = st_coordinates(centroid)[, 1],
    y = st_coordinates(centroid)[, 2],
    label = toupper(ADM1_PT)
  )

# Top 3 worst districts per province (for labeling)
sf_top3 <- sf_exposure |>
  filter(ADM1_PT %in% c("Gaza", "Maputo", "Sofala")) |>
  group_by(valid_date, ADM1_PT) |>
  slice_max(sum, n = 3) |>
  ungroup() |>

  mutate(centroid = st_centroid(geom)) |>
  mutate(
    x = st_coordinates(centroid)[, 1],
    y = st_coordinates(centroid)[, 2]
  )

ggplot() +
  geom_sf(data = sf_exposure, aes(fill = sum), color = "white", linewidth = 0.3) +
  geom_sf(data = sf_provinces, fill = NA, color = "grey70", linewidth = 0.5) +
  geom_sf(data = sf_highlight, fill = NA, color = "#D62828", linewidth = 1.5) +
  geom_text(
    data = sf_prov_labels,
    aes(x = x, y = y, label = label),
    size = 5, color = "grey50", fontface = "bold", alpha = 0.7
  ) +
  geom_label_repel(
    data = sf_top3,
    aes(x = x, y = y, label = ADM2_PT),
    size = 4,
    fill = alpha("white", 0.7),
    label.size = 0.3,
    fontface = "bold",
    segment.color = "grey40",
    segment.size = 0.4,
    box.padding = 0.5,
    point.padding = 0.3,
    max.overlaps = 20,
    min.segment.length = 0
  ) +
  scale_fill_viridis_c(
    option = "inferno",
    direction = -1,
    labels = scales::comma,
    na.value = "grey90",
    name = "Population Exposed"
  ) +
  facet_wrap(~valid_date, ncol = 1) +
  labs(
    title = "FloodScan Exposure by District",
    subtitle = "Gaza, Maputo, and Sofala highlighted | Top 3 districts labeled per province",
    caption = "Data: FloodScan & WorldPop"
  ) +
  theme_void(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 18, margin = margin(b = 10)),
    legend.position = "bottom",
    legend.key.width = unit(4, "cm"),
    legend.key.height = unit(0.6, "cm"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(color = "grey40", size = 14, margin = margin(b = 20))
  )
```

